Your time series is (probably) too short

### 2017-10-01 writing/manuscript_draft.Rmd
Version submitted to Biological Conservation and the second version to PeerJ preprints

output:
  pdf_document:
    fig_caption: yes
    keep_tex: yes
    number_sections: yes
  word_document: default
---

Title: Minimum time required to detect population trends: the need for long-term monitoring programs
\vspace{7 mm}

@@

$^2$Corresponding author: eawhite@ucdavis.edu <!--, 1 480 203 7931-->
\vspace{3 mm}

Number of words (including 300 words per figure):  5900 
\vspace{1 mm}

@@

@@

<!--To be submitted to: \emph{Biological Conservation}
\vspace{3 mm}

Available as a preprint at: www.peerj.com
\vspace{3 mm}

Color does not need to be used for any figures in print
\vspace{3 mm}-->

@@

\pagebreak

\begin{abstract}
Long-term time series are necessary to better understand population dynamics, assess species' conservation status, and make management decisions. However, population data are often expensive, requiring a lot of time and resources. When is a population time series long enough to address a question of interest? We determine the minimum time series length required to detect significant increases or decreases in population abundance. To address this question, we use simulation methods and examine 878 populations of vertebrate species. Here we show that 15-20 years of continuous monitoring are required in order to achieve a high level of statistical power. For both simulations and the time series data, the minimum time required depends on trend strength, population variability, and temporal autocorrelation. These results point to the importance of sampling populations over long periods of time. We argue that statistical power needs to be considered in monitoring program design and evaluation. Time series less than 15-20 years are likely underpowered and potentially misleading.
\end{abstract}

@@

How long of a time series is actually necessary? This question has important implications for both research and management [@Nichols2006]. Scientists need to know the time series length required to address a specific question. A short time series may lead to wrong conclusions given large natural year-to-year variability. Managers need to know when action is needed for a population. Therefore, they must understand when population trend over time is actually meaningful. The IUCN Red List Categories and Criteria suggest, under Criterion A2, a species qualifies as vulnerable if it has experienced a 30% decline over 10 years, or 3 generations [@IUCN2012]. For both scientific and management questions, because sampling is typically expensive, we also do not want to sample for longer than is necessary. For example, @Gerber1999 investigated the minimum time series required to estimate population growth of the endangered, but recovering, eastern North Pacific gray whale (\emph{Eschrichtius robustus}). They used a long-term census to retroactively determine the minimum time series required to assess threat status. They found that only 11 years were needed, eight years before the delisting decision was made. This highlights the importance of estimating the minimum time series required as an earlier decision would have saved time and money [@Gerber1999].

@@

For time series data, two general approaches to estimating sample size are appropriate. <!--Typically, calculations of sample size are given by a formula specific to standard statistical tests (e.g. t-test). However, many ecological scenarios require the use of simulations [@Bolker2008; @Johnson2015]. As a first approach,-->Simulations can be designed for a specific population and question [@Bolker2008; @Johnson2015]. Simple models can be simulated with parameter values corresponding to a population of interest [@Gerrodette1987]. Statistical power can then be calculated as the fraction of simulations that meet some criteria. The specific criteria depend on the question at hand. For example, given a time series, when is the slope from linear regression significantly different from zero? In other words, when is the time series significantly increasing or decreasing? It is then possible to determine how power changes with a variable of interest. For example, time series can be simulated for different lengths of time. From these simulations, the minimum time series length required is calculated to meet certain levels of statistical significance and power [@Bolker2008].

In addition to using simulations, empirical time series can also be used. Multiple replicates of similar populations are usually not available, but it is possible to subsample an empirical time series [@Gerber1999]. Subsamples of different lengths can then be evaluated to see which fraction of subsamples meet some criteria, again a measure of statistical power. Similar to the simulation approach, these measures of power can be used to determine the minimum time series required for a particular question of interest.

Past work has investigated questions related to the minimum time series required to estimate trends in population size over time [@Wagner2009; @Giron-Nava2017]. For example, @Rhodes2011 examined the optimal allocation of effort between spatial and temporal replicates. Using simple populations models, they found that the allocation of effort depends on environmental variation, spatial and temporal autocorrelation, and observer error. @Rueda-Cediel2015 also used a modeling approach, but parameterized a model specifically for a threatened snail, \emph{Tasmaphena lamproides}. They found that for this short-lived organism, 15 years was adequate to assess long-term trends in abundance. @Hatch2003 examined seabird populations and determined the minimum time required to achieve high statistical power ranged between 11 and 69 years. However, these studies, and other past work, have typically focused either on theoretical aspects of monitoring design or focused on only a few species.

@@

Statistical power is then the fraction of simulations that meet some criteria. Here, our criteria is whether the slope parameter from linear regression is significant at the 0.05 threshold. Statistical power of 0.8 would indicate that, if there was a true trend in abundance, we would have a 0.8 probability of detecting the trend. We also tested the effect of varying both the significance level and statistical power (Fig. \ref{fig:min_time_vs_alpha_beta}).

@@

## Data source
We use a database of `r length(table(dat$ID))` population time series compiled in [@Keith2015]; they compared the predictability of growth rates among populations. The data are originally from the Global Population Dynamics Database [@GPDD2010]. We filtered out short time series (less than `r pop_info$series_length[1]` years), and those with missing data, and were left with `r nrow(pop_info)` time series. The data includes information on `r length(table(pop_info$Binomial))` vertebrate species with a focus on mammals, birds, and fish. The data also includes information on generation length and census specifications. For each time series, we also calculated variables of interest like variance in population size, long-term trend in abundance (slope coefficient from simple linear regression), and temporal autocorrelation. All analyses were conducted in R [@RCoreTeam2016]. <!--Lastly, we also calculated the minimum number of years required (10 years or 3 generations) under IUCN Red List Criterion A2 to classify a species as vulnerable [@IUCN2012].-->

@@

We assume that each time series in our database is long enough to include all necessary information (e.g. variability) about the population. In other words, each time series is a representative sample. We first take all possible contiguous subsamples of each time series. For example, a time series of 35 years would have 34 possible contiguous subsamples of length 2, 34 possible contiguous subsamples of length 3, and continuing until 1 possible contiguous subsample of length 35 [@Gerber1999; @Giron-Nava2017]. Next, we run a linear regression for each subsampled time series. Then, we determine the fraction of subsamples of a particular length that have estimated slope coefficients which are statistically different from zero. We only look at the fraction of samples where the long-term, or ``true'', time series also has a significant slope. This fraction is a measure of statistical power. Lastly, we determine which subsample length is required to achieve a certain threshold of statistical power (0.8, @Cohen1992). The subsampled length meeting these criteria is the minimum time series length required ($T_{min}$).

@@

We chose a simple model, but any other population model could be used (see example in Fig. \ref{fig:min_time_dist_log_pop}). Ideally, the specific model choice should be tailored to the population of interest. We explored how the simulation approach can be applied to more biologically-realistic population models (Fig. \ref{fig:shark_example}). More specifically, we determined the minimum time required to estimate long-term population trends using a stochastic, age-structured model of lemon shark population dynamics in the Bahamas [@White2014]. We found that over 27 years of continuous monitoring were needed in this particular scenario (Fig. \ref{fig:shark_example}). Similar to the simulation approach described above, the minimum time required for the lemon shark population was strongly dependent on model parameters.

@@

```{r,echo=F,fig.cap='Distributions of the minimum time required for four different biological classes. The minimum time required calculation corresponds to a significance level of 0.05 and statistical power of 0.8.\\label{fig:min_time_dist}',fig.width=6}
# Pull in analysis of average min time needed for **empirical species** (save analysis as Rdata file or something)
 load("../cleaned-data/cleaned_timeseries_database_with_min_time_linear_regression.Rdata")

@@

In line with theoretical predictions [@Rhodes2011], we also found $T_{min}$ was strongly correlated with the trend strength, variability in population size, and temporal autocorrelation (Fig. \ref{fig:correlates}). Contrary to our prior expectations, we also found that $T_{min}$ did not correlate with any biological variables of interest (Fig. \ref{fig:biological_correlates}). We initially hypothesized that species with longer lifespans or generation times may require a longer sampling period. Our result could have been a result of at least two factors. First, the data we used may not include a diverse enough set of species with different life history traits. Second, the question we poised, whether a population is increasing or decreasing, was specifically concerned with trends in population density over time. Therefore, life-history characteristics may be more important for other questions, like estimating species extinction risk [@Hutchings2012].

@@

An important related question, is the optimal allocation of sampling effort in space versus time. In a theoretical investigation of this question, @Rhodes2011 found that the optimal allocation of sampling depended strongly on temporal and spatial autocorrelation. If spatial population dynamics were highly correlated, then it was better to sample more temporally, and vice versa. Our work supports this idea as populations with strong temporal autocorrelation needed less years of sampling (Fig. \ref{fig:correlates}). @Morrison2008 also studied the optimal allocation of sampling effort in space versus time, but used emprical data from a long-term census of the collared pika (\emph{Ochotona collaris}) found in the Yukon. They estimated long-term growth rates among three subpopulations over a 10-year period. They found that censuses less than 5 years may be misleading and that extrapolating from one population to another, even when nearby geographically, may be untenable.

@Seavy2007 asked whether statistical power was even a useful framework for assessing long-term population trends. They used 24 years of census data on Red-tailed Tropicbirds (\emph{Phaethon rubricauda}) in Hawaii and showed that to detect a 50% decline over 10 years almost always resulting in high statistical power (above 0.8). Therefore, they cautioned against only using power analyses to design monitoring schemes and instead argued for metrics that would increase precision. For example, @Seavy2007 suggest improving randomization, reducing bias, and increases detection probability when designing and evaluating monitoring programs. We agree that power analyses should not be the only consideration when designing monitoring schemes. However, unlike @Seavy2007, our results indicate that longer than 10 years is often needed to achieve high statistical power. Therefore, the design of monitoring programs should include calculations of statistical power, the allocation of sampling in space versus time [@Rhodes2011], and metrics to increase precision. Ideally, a formal decision analysis to evaluate these different factors would be conducted to design or assess any monitoring program [@Hauser2006;@McDonald-Madden2010].

@@

Our work has some limitations in determining the minimum time series length required. First, $T_{min}$ is particular to the specific question of interest. An additional complication is that for our empirical approach, the subsampling of the full time series allows for estimates of power, but the individual subsamples are clearly not independent of one another. Further, estimates of $T_{min}$ depend on chosen values of $\alpha$ and $\beta$ (Fig. \ref{fig:min_time_vs_alpha_beta}). In an ideal setting, we would build a specific population model parameterized for each population of interest. Then, model simulations could be used to estimate the minimum time series required to address each specific question of interest. Clearly, this is not always practical, especially if conducting analyses for a wide array of species as we do here. In addition, our statistical models suggest that $T_{min}$ does not correlate with any biological variables of interest, at least for the question of linear regression (Fig. \ref{fig:biological_correlates}). Therefore, it is not possible to use these results to predict $T_{min}$ for another population, even if the population is of a species with a similar life-history to one in our database.

@@

We use a database of `r nrow(pop_info)` populations to determine the minimum time series length required to detect population trends. This goes beyond previous work that either focused on theoretical investigations or a limited number of species. We show that to identify long-term changes in abundance, 15-20 years of continuous monitoring are often required (Fig. \ref{fig:min_time_dist}). In line with theoretical predictions [@Gerrodette1987], we also show that $T_{min}$ is strongly correlated with the long-term population trend (i.e. rate of increase), variability in population size, and the temporal autocorrelation (Fig. \ref{fig:correlates}). Our work implies that for many populations, time series less than 15-20 years are probably not reliable for detecting population trends. This result highlights the importance of long-term monitoring programs. From both a scientific and management perspective estimates of $T_{min}$ are important. If a time series is too short, we lack statistical power to reliably detect long-term population trends. In addition, a time series that is too long may be a poor use of already limited funds [@Gerber1999]. Future work should examine other species, with a wider range of life history characteristics. In addition, similar approaches can be used to determine the minimum time series length required to address additional questions of interest.

>>>>

output:
  pdf_document:
    fig_caption: yes
    keep_tex: yes
    number_sections: yes
  word_document: default
---
<!--
Title: Minimum time required to detect population trends: the need for long-term monitoring programs
\vspace{7 mm}

@@

$^2$Corresponding author: eawhite@ucdavis.edu , 1 480 203 7931
\vspace{3 mm}

Number of words (including 300 words per figure):  5950 
\vspace{1 mm}

@@

Number of references: 32
\vspace{1 mm}

@@

To be submitted to: \emph{Biological Conservation}
\vspace{3 mm}

Available as a preprint at: https://peerj.com/preprints/3168/
\vspace{3 mm}

Color does not need to be used for any figures in print
\vspace{3 mm}

@@

\pagebreak-->

 \textbf{Minimum time required to detect population trends: the need for long-term monitoring programs}
 \vspace{3 mm}

Easton R. White$^{*}$

\begin{footnotesize} Center for Population Biology, University of California, Davis, California 95616 USA

*eawhite@ucdavis.edu
\end{footnotesize}

\vspace{5 mm}
\linenumbers

\begin{abstract}
Long-term time series are necessary to better understand population dynamics, assess species' conservation status, and make management decisions. However, population data are often expensive, requiring a lot of time and resources. When is a population time series long enough to address a question of interest? We determine the minimum time series length required to detect significant increases or decreases in population abundance. To address this question, we use simulation methods and examine 878 populations of vertebrate species. Here we show that 15-20 years of continuous monitoring are required in order to achieve a high level of statistical power. For both simulations and the time series data, the minimum time required depends on trend strength, population variability, and temporal autocorrelation. These results point to the importance of sampling populations over long periods of time. We argue that statistical power needs to be considered in monitoring program design and evaluation. Time series less than 15-20 years are likely under-powered and potentially misleading.
\end{abstract}

@@

How long of a time series is actually necessary? This question has important implications for both research and management [@Nichols2006]. Scientists need to know the time series length required to address a specific question. A short time series may lead to wrong conclusions given large natural year-to-year variability [@McCain2016]. Managers need to know when action is needed for a population. Therefore, managers must understand when population trend over time is actually meaningful. For example, the IUCN Red List Categories and Criteria suggest, under Criterion A2, a species qualifies as vulnerable if it has experienced a 30% decline over 10 years, or 3 generations [@IUCN2012]. For both scientific and management questions, because sampling is typically expensive, we also do not want to sample for longer than is necessary. For example, @Gerber1999 investigated the minimum time series required to estimate population growth of the endangered, but recovering, eastern North Pacific gray whale (\emph{Eschrichtius robustus}). They used a long-term census to retroactively determine the minimum time series required to assess threat status. They found that only 11 years were needed, eight years before the delisting decision was made. This highlights the importance of estimating the minimum time series required as an earlier decision would have saved time and money [@Gerber1999].

@@

For time series data, two general approaches to estimating sample size are appropriate. <!--Typically, calculations of sample size are given by a formula specific to standard statistical tests (e.g. t-test). However, many ecological scenarios require the use of simulations [@Bolker2008; @Johnson2015]. As a first approach,-->Simulations can be designed for a specific population and question [@Bolker2008; @Johnson2015]. Simple models can be simulated with parameter values corresponding to a population of interest [@Gerrodette1987]. Statistical power is the proportion of simulations that meet some set of criteria. The specific criteria depend on the question at hand. For example, given a time series, when is the slope from linear regression significantly different from zero? In other words, when is the time series significantly increasing or decreasing? It is then possible to determine how power changes with a variable of interest. For example, time series can be simulated for different lengths of time. From these simulations, the minimum time series length required to meet certain levels of statistical significance and power is estimated [@Bolker2008].

In addition to using simulations, empirical time series can also be used. Multiple replicates of similar populations are usually not available, but it is possible to subsample an empirical time series [@Gerber1999; @Brashares2005]. Subsamples of different lengths can then be evaluated to estimate the proportion of subsamples meeting some criteria, again a measure of statistical power. Similar to the simulation approach, this measures of power can be used to determine the minimum time series required for a particular question of interest.

Past work has investigated questions related to the minimum time series required to estimate trends in population size over time [@Wagner2009; @Giron-Nava2017]. For example, @Rhodes2011 examined the optimal allocation of effort between spatial and temporal replicates. Using simple populations models, they found that the allocation of effort depends on environmental variation, spatial and temporal autocorrelation, and observer error. @Rueda-Cediel2015 also used a modeling approach, but parameterized a model specific for a threatened snail, \emph{Tasmaphena lamproides}. They found that for this short-lived organism, 15 years was adequate to assess long-term trends in abundance. @Hatch2003 examined seabird populations and determined the minimum time required to achieve high statistical power ranged between 11 and 69 years. However, these studies, and other past work, have typically focused either on theoretical aspects of monitoring design or focused on only a few species.

@@

Statistical power is then the proportion of simulations that meet some criteria. Here, our criteria is whether the slope parameter from linear regression is significant at the 0.05 threshold with statistical power of 0.8. Statistical power of 0.8 would indicate that, if there was a true trend in abundance, we would have a 0.8 probability of detecting the trend. We also tested the effect of varying both the significance level and statistical power (Fig. \ref{fig:min_time_vs_alpha_beta}).

@@

## Data source
We use a database of `r length(table(dat$ID))` population time series compiled in [@Keith2015]; they compared the predictability of growth rates among populations. The data are originally from the Global Population Dynamics Database [@GPDD2010]. We filtered out short time series (less than `r pop_info$series_length[1]` years), and those with missing data, and were left with `r nrow(pop_info)` time series. The data includes information on `r length(table(pop_info$Binomial))` vertebrate species with a focus on mammals, birds, and fish. The data also includes information on generation length and survey specifications. For each time series, we also calculated variables of interest like variance in population size, long-term trend in abundance (slope coefficient from simple linear regression), and temporal autocorrelation. All analyses were conducted in R [@RCoreTeam2016]. <!--Lastly, we also calculated the minimum number of years required (10 years or 3 generations) under IUCN Red List Criterion A2 to classify a species as vulnerable [@IUCN2012].-->

@@

We assume that each time series in our database is long enough to include all necessary information (e.g. variability) about the population. In other words, each time series is a representative sample. We first take all possible contiguous subsamples of each time series. For example, a time series of 35 years would have 34 possible contiguous subsamples of length 2, 34 possible contiguous subsamples of length 3, and continuing until 1 possible contiguous subsample of length 35 [@Gerber1999; @Giron-Nava2017]. Next, we run a linear regression for each subsampled time series. Then, we determine the proportion of subsamples of a particular length that have estimated slope coefficients which are statistically different from zero. We only look at the proportion of samples where the long-term, or ``true'', time series also has a significant slope. This proportion is a measure of statistical power. Lastly, we determine which subsample length is required to achieve a certain threshold of statistical power (0.8, @Cohen1992). The minimum subsampled length that met these criteria is the minimum time series length required ($T_{min}$).

@@

We chose a simple model, but any other population model could be used (see example in Fig. \ref{fig:min_time_dist_log_pop}). Ideally, the specific model choice should be tailored to the population of interest. We explored how the simulation approach could be applied to more biologically-realistic population models (Fig. \ref{fig:shark_example}). More specifically, we determined the minimum time required to estimate long-term population trends using a stochastic, age-structured model of lemon shark population dynamics in the Bahamas [@White2014]. We found that over 27 years of continuous monitoring were needed in this particular scenario (Fig. \ref{fig:shark_example}). Similar to the simulation approach described above, the minimum time required for the lemon shark population was strongly dependent on model parameters.

@@

```{r,echo=F,fig.cap='Distributions of the minimum time required for populations from four different biological classes. The minimum time required calculation corresponds to a significance level of 0.05 and statistical power of 0.8.\\label{fig:min_time_dist}',fig.width=6}
# Pull in analysis of average min time needed for **empirical species** (save analysis as Rdata file or something)
 load("../cleaned-data/cleaned_timeseries_database_with_min_time_linear_regression.Rdata")

@@

In line with theoretical predictions [@Rhodes2011], we also found $T_{min}$ was strongly correlated with the trend strength, variability in population size, and temporal autocorrelation (Fig. \ref{fig:correlates}). Contrary to our prior expectations, we also found that $T_{min}$ did not correlate with any biological variables of interest (Fig. \ref{fig:biological_correlates}). We initially hypothesized that species with longer lifespans or generation times may require a longer sampling period. Our result could have been a result of at least two factors. First, the data we used may not include a diverse enough set of species with different life history traits. Second, the question we poised, whether a population is increasing or decreasing, was specifically concerned with trends in population density over time. Therefore, life-history characteristics may be more important for other questions, like estimating species extinction risk [@Hutchings2012]. For example, @Blanchard2007 used detailed simulations of spatially-distributed fisheries to compare survey designs. They found that statistical power depended on survey design, temperature preferences and the degree of population patchiness.

@@

An important related question, is the optimal allocation of sampling effort in space versus time. In a theoretical investigation of this question, @Rhodes2011 found that the optimal allocation of sampling depended strongly on temporal and spatial autocorrelation. If spatial population dynamics were highly correlated, then it was better to sample more temporally, and vice versa. Our work supports this idea as populations with strong temporal autocorrelation needed less years of sampling (Fig. \ref{fig:correlates}). @Morrison2008 also studied the optimal allocation of sampling effort in space versus time, but used emprical data from a long-term survey of the collared pika (\emph{Ochotona collaris}) found in the Yukon. They estimated long-term growth rates among three subpopulations over a 10-year period. They found that surveys less than 5 years may be misleading and that extrapolating from one population to another, even when nearby geographically, may be untenable.

@Seavy2007 asked whether statistical power was even a useful framework for assessing long-term population trends. They used 24 years of census data on Red-tailed Tropicbirds (\emph{Phaethon rubricauda}) in Hawaii and showed that to detect a 50% decline over 10 years almost always resulting in high statistical power (above 0.8). Therefore, they cautioned against only using power analyses to design monitoring schemes and instead argued for metrics that would increase precision. For example, @Seavy2007 suggest improving randomization, reducing bias, and increases detection probability when designing and evaluating monitoring programs. We agree that power analyses should not be the only consideration when designing monitoring schemes. However, unlike @Seavy2007, our results indicate that longer than 10 years is often needed to achieve high statistical power.

The design of monitoring programs should include calculations of statistical power, the allocation of sampling in space versus time [@Rhodes2011], and metrics to increase precision. Ideally, a formal decision analysis to evaluate these different factors would be conducted to design or assess any monitoring program [@Hauser2006; @McDonald-Madden2010]. This type of formal decision analysis would also include information on the costs of monitoring. These costs include the actual costs of sampling [@Brashares2005] and the ecological costs on inaction [@Thompson2000].

@@

Our work has some limitations in determining the minimum time series length required. First, $T_{min}$ is particular to the specific question of interest. An additional complication is that for our empirical approach, the subsampling of the full time series allows for estimates of power, but the individual subsamples are clearly not independent of one another. Further, estimates of $T_{min}$ depend on chosen values of $\alpha$ and $\beta$ (Fig. \ref{fig:min_time_vs_alpha_beta}). In an ideal setting, we would build a specific population model parameterized for each population of interest [@McCain2016]. Then, model simulations could be used to estimate the minimum time series required to address each specific question of interest. Clearly, this is not always practical, especially if conducting analyses for a wide array of species as we do here. In addition, our statistical models suggest that $T_{min}$ does not correlate with any biological variables of interest, at least for the question of linear regression (Fig. \ref{fig:biological_correlates}). Therefore, it is not possible to use these results to predict $T_{min}$ for another population, even if the population is of a species with a similar life-history to one in our database.

@@

We use a database of `r nrow(pop_info)` populations to determine the minimum time series length required to detect population trends. This goes beyond previous work that either focused on theoretical investigations or a limited number of species. We show that to identify long-term changes in abundance, 15-20 years of continuous monitoring are often required (Fig. \ref{fig:min_time_dist}). In line with theoretical predictions [@Gerrodette1987], we also show that $T_{min}$ is strongly correlated with the long-term population trend (i.e. rate of increase), variability in population size, and the temporal autocorrelation (Fig. \ref{fig:correlates}).

Our work implies that for many populations, time series less than 15-20 years are probably not reliable for detecting population trends. This result highlights the importance of long-term monitoring programs. From both a scientific and management perspective estimates of $T_{min}$ are important. If a time series is too short, we lack statistical power to reliably detect long-term population trends. In addition, a time series that is too long may be a poor use of already limited funds [@Gerber1999]. However, short time series data can not be used as an excuse for inaction. For example, @Thompson2000 demonstrate that long-term studies may be required to detect population trends for certain species. When a population trend is detected, it may be too late for management action. In these situations, the precautionary principle may be more appropriate [@Thompson2000]. Future work should examine other species, with a wider range of life history characteristics. In addition, similar approaches can be used to determine the minimum time series length required to address additional questions of interest.
